---
title: "APCA_monitors"
author: "Rachel Tao"
date: "9/3/2021"
output: github_document
---

```{r setup, include=FALSE}
library(tidyverse) 
library(lubridate) 
library(ggmap) 
library(ggcorrplot) 
library(ggplot2) 
library(gplots)
library(ggfortify) 
library(ggrepel) 
library(gridExtra) 
library(GPArotation) 
library(GGally) 
library(knitr) 
library(patchwork) 
library(psych) 
library(factoextra) 
library(RColorBrewer) 
library(kableExtra) 
library(pals)
library(httr)

knitr::opts_chunk$set(
	echo = FALSE,
	fig.asp = 0.6,
	fig.width = 6,
	warning = FALSE,
	out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_brewer(palette = "Spectral")
scale_fill_discrete = scale_fill_brewer(palette = "Spectral")

```

<<<<<<< HEAD
```{r}
# Load in total PM data

pm25_tot <- read_csv("./data/pm2.5_total_components_daily_nyc_avg.csv") %>% 
  rename(
    date = date_local,
    pm25 = pm2.5_total
  ) %>% 
  select(date, pm25)

pm25_tot_raw <- read_csv("./data/pm2.5_nyc_master.csv") %>% 
  filter(site_num %in% c(110, 62, 134, 124))

pm25_tot_raw %>% 
  filter(site_num == 110) %>% 
  select(poc, date_local) %>% 
  group_by(poc) %>% 
  summarize(max(date_local), min(date_local))

pm25_tot_raw %>% select(site_num, poc) %>% unique()

pm25_tot_raw %>% 
  filter(
    site_num == 110,
    poc == 1
    ) %>% 
  select(date_local) %>% 
  view()

pm25_tot_raw %>% 
  filter(site_num == 62) %>% 
  select(poc, date_local) %>% 
  group_by(poc) %>% 
  summarize(max(date_local), min(date_local))

pm25_tot_raw %>% 
  filter(site_num == 134) %>% 
  select(poc, date_local) %>% 
  group_by(poc) %>% 
  summarize(max(date_local), min(date_local))


```


```{r echo=FALSE, message=FALSE, warning=FALSE}

# Load in speciated PM data

pm25_tot <- read_csv("./data/pm2.5_total_components_daily_nyc_avg.csv") %>% 
  rename(
    date = date_local,
    pm25 = pm2.5_total
  ) %>% 
  select(date, pm25)

all_elements_bx <- read_csv("./data/bronx_daily_pm2.5_components.csv") %>% 
  janitor::clean_names() %>% 
  rename(ammonium = ammonium_ion,
         EC = elemental_carbon,
         OC = organic_carbon,
         potassium = potassium_ion,
         nitrate = total_nitrate) %>% 
  filter(date > "2006-12-31",
         date < "2016-01-01")

conc <- all_elements %>% 
  select(-antimony, -cobalt) %>% 
  left_join(pm25_tot, by = "date") %>% 
  relocate(date, pm25)

# multiply all concentrations by 1000 to convert to ng/m3
i <- c(2:28)
conc[ , i] <- apply(conc[ , i], 2,
                    function(x)
                      x*1000)

missing_element <- sapply(conc, function(x) sum(is.na(x))/nrow(conc))

dates <- tibble(seq(as.Date('2007-01-01'), as.Date('2015-12-31'), by = "days")) %>% 
  rename(date = 'seq(as.Date("2007-01-01"), as.Date("2015-12-31"), by = "days")')

time <- 
  left_join(dates, conc, by = "date")

element_scatter <- 
  conc %>% 
  ggplot(aes(x = date, y = cadmium)) +
  geom_point()

```

```{r}
# correlation matrix

ggcorr(all_elements, method = c("pairwise.complete.obs", "pearson"), limits = FALSE,
       label = TRUE, label_size = 1.5, label_alpha = TRUE,
       hjust = 1, size = 3, color = "grey50", layout.exp = 4) + ggtitle("Correlation Matrix: NYC Air Pollution Dataset")

```


```{r eval=FALSE, include=FALSE}
# which days have unusually high levels of potassium?

K_year <- conc %>% 
  mutate(year = format(date, "%Y"),
         CommonDate = as.Date(paste0("2009", format(date, "%j")), "%Y-%j")) %>%
  ggplot(aes(x = CommonDate, y = potassium)) +
  geom_line() +
  facet_grid(facets = year ~., margins = FALSE) +
  scale_x_date(
    labels = function(x) format(x, "%d-%b"),
    breaks = "month") +
  xlab("") +
  ggtitle("Potassium concentrations by year")

K_year

conc %>% 
  select(date, potassium) %>% 
  filter(potassium > 200) %>% 
  kable(caption = "dates  where Potassium > 200 ng/m3")

conc %>% 
  select(date, potassium) %>% 
  filter(potassium > 150) %>% 
  kable(caption = "dates  where Potassium > 150 ng/m3")

# we will exclude the days surrounding 4th of july, which consistently have
# unusually high potassium levels due to fireworks.

#try the same thing with cadmium

Cd_year <- conc %>% 
  ggplot(aes(x = date, y = cadmium)) +
  geom_line() +
  xlab("") +
  ggtitle("Cadmium concentrations by year")


```


```{r}
# finish cleaning data using information about fireworks

fireworks <- as_date(c(
                     ymd("2007-07-02"):ymd("2007-07-06"),
                     ymd("2008-07-02"):ymd("2008-07-06"),
                     ymd("2009-07-02"):ymd("2009-07-06"),
                     ymd("2010-07-02"):ymd("2010-07-06"),
                     ymd("2011-07-02"):ymd("2011-07-06"),
                     ymd("2012-07-02"):ymd("2012-07-06"),
                     ymd("2013-07-02"):ymd("2013-07-06"),
                     ymd("2014-07-02"):ymd("2014-07-06"),
                     ymd("2015-07-02"):ymd("2015-07-06")))

conc <- conc %>%
  filter(!(date %in% fireworks )) #filter out 4th of July fireworks!

```


```{r echo=FALSE, message=FALSE, warning=FALSE}
# Means, correlations and variances

conc_corr <- 
  conc %>% 
  select(-date)

corr <- 
  summarize(conc_corr, corr = cor(conc_corr, use = "pairwise.complete.obs"))

conc_summarize <- 
  conc_corr %>% 
  pivot_longer(everything(),
               names_to = "element",
               values_to = "concentration") %>% 
  group_by(element)

means <- conc_summarize %>% 
  summarize(mean = mean(concentration, na.rm = TRUE)) %>% 
  arrange(element)

vars <- conc_summarize %>% 
  summarize(variance = var(concentration, na.rm = TRUE)) %>% 
  arrange(element)

```

```{r PROVIDE FACTORS, echo=FALSE, message=FALSE, warning=FALSE}
# Scaled factor analysis - must provide number of factors here

# set mean = 0, SD = 1
zspec_pm25 <- 
  conc %>% 
  select(-date, -pm25) %>% 
  scale(center = TRUE, scale = TRUE)

# factor analysis - must provide number of factors!
factors <- 
  fa(zspec_pm25,
     nfactors = 6, max.iter = 500,
     rotate = "varimax",
     scores = "regression",
     SMC  = TRUE,
     fm = "pa")

# calculate factor scores, loadings, and weights (aka standardized scoring coefficients)
fact <- as_tibble(factors$scores)
loadings <- factors$loadings
factor_stat <- factors$weights %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column("element") %>% 
  arrange(element)

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# calculate absolute factor scores for each day (Equations 5-7 in statistical methods of Kavouras 2001)

# join standardized scoring coefficients with means and variances
coeff <- left_join(means, vars, by = "element") %>% 
  left_join(factor_stat, by = "element")

# calculate minimum values to find absolute factor scores (equations 5 and 6)
min <- coeff %>% 
  pivot_longer(contains("PA"),
               names_to = "factor_num",
               names_prefix = "PA",
               values_to = "factor_coeff") %>% 
  mutate(minimum = factor_coeff*(-mean/(variance^(1/2)))) %>% 
  select(element, factor_num, minimum) %>% 
  group_by(factor_num) %>% 
  summarize(sum = sum(minimum, na.rm = TRUE)) %>% 
  pivot_wider(factor_num:sum,
              names_from = factor_num,
              names_prefix = "minimum_",
              values_from = sum)

# reattach factor scores to dates and calculate absolute factor scores for each day
# (equation 7)
a <- conc %>% 
  select(date) %>% 
  cbind(fact) %>% 
  pivot_longer(contains("PA"),
               names_to = "factor_num",
               names_prefix = "PA",
               values_to = "factor_score") %>% 
  mutate(factor_num = as.numeric(factor_num)) %>% 
  cbind(min) %>% 
  pivot_longer(contains("minimum"),
               names_to = "min_num",
               names_prefix = "minimum_",
               values_to = "minimum") %>% 
  mutate(min_num = as.numeric(min_num)) %>% 
  filter(factor_num == min_num) %>% 
  mutate(factor_score = factor_score - minimum) %>% 
  select(-min_num, -minimum)

```


```{r message=FALSE, warning=FALSE}
# factor means

# this is a check - not necessary for next steps
factor_means <- a %>% 
  select(factor_num, factor_score) %>% 
  group_by(factor_num) %>% 
  summarize(mean_score = mean(factor_score, na.rm = TRUE))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# Regress each element on factors to see source contribution

# define 'regress' as regression function with y as dependent variable and x as independent variable(s)
# return tidy output for beta values and r-squared value
regress = function(y) {
  
  z = lm(y ~ x)
  
  r.squared = broom::glance(z) %>% select(r.squared)

  betas = broom::tidy(z) %>% pivot_wider(term:estimate,
                                         names_from = "term",
                                         values_from = "estimate")
  
  return(cbind(r.squared, betas))
  
}

# independent variables for regression will be factors from factor analysis. define these as 'x'
x <- a %>% 
  pivot_wider(everything(),
              names_from = factor_num,
              names_prefix = "factor_",
              values_from = factor_score) %>% 
  select(contains("factor")) %>% as.matrix()

# dependent variable for regression will be actual element concentration. define this as y.
y <- conc %>% select(-date)

# repeat 'regress' for each element using 'x' as factors and 'y' as concentrations)
regress <- map(y, regress) %>% 
  enframe(name = "element") %>% 
  unnest(c(value)) %>% 
  janitor::clean_names() %>% 
  rename_at(vars(contains("factor")), funs(str_replace(., "xfactor", "beta")))

# dataset storing r-squared values
r <- regress %>% 
  select(element, r_squared)

# dataset storing beta values
betas <- regress %>% 
  select(element, starts_with("beta")) %>% 
  nest(betas = starts_with("beta"))

```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# calculate source contributions

# combine betas with factor scores in a list 
sources = lapply(betas$betas, cbind, a)

# in the above step, element names are automatically removed. rename list elements:
names(sources) <- betas$element

# calculate source contributions by multiplying absolute factor scores by beta values
sources <- lapply(sources, as_tibble) %>% 
  bind_rows(.id = "element") %>%
  pivot_longer(starts_with("beta"),
               names_to = "beta_num",
               names_prefix = "beta_",
               values_to = "beta_value") %>% 
  mutate(
    beta_num = as.numeric(beta_num)
  ) %>% 
  filter(factor_num == beta_num) %>% 
  mutate(
    srce = factor_score*beta_value
  ) %>% 
  mutate(srce_num = as.character(beta_num)) %>% 
  select(element, date, srce_num, srce) 

conc_long <- conc %>% 
  pivot_longer(!date,
               names_to = "element",
               values_to = "concentration") %>% 
  arrange(element)

source_conc <- left_join(sources, conc_long, by = c('element', 'date'))

source_conc %>% 
  filter(element == "pm25") %>% 
  group_by(srce_num) %>% 
  summarize(sd = sd(srce, na.rm = TRUE),
            mean = mean(srce, na.rm = TRUE))

# calculate mean source contributions for each element
mean_source <- source_conc %>% 
  group_by(element, srce_num) %>% 
  summarize(source_mean = mean(srce, na.rm = TRUE),
            MeanConc = mean(concentration, na.rm = TRUE)) %>% 
  pivot_wider(names_from = "srce_num",
              names_prefix = "source_",
              values_from = "source_mean")

```

## Loadings
```{r echo=FALSE, message=FALSE, warning=FALSE}
# Source contributions to each component of PM2.5

# calculate mean predicted concentration and percent error
SA_Varimax <- mean_source %>% 
  ungroup(element) %>% 
  left_join(r, by = "element") %>% 
  mutate(
    PredConc = reduce(select(., contains("source")), `+`),
    Pct_error = ((PredConc - MeanConc)/MeanConc)*100
  )

i <- c(2:11)
SA_Varimax[ , i] <- apply(SA_Varimax[ , i], 2,
                    function(x)
                      round(x, digits = 2))

#write_csv(SA_Varimax, "./experimental_results/SA_table.csv")

SA_Varimax %>% kable(booktabs = T, align = "c") %>% 
  kable_styling(full_width = T)

# calculate proportion of each element coming from each source

SA_proportion_full <- source_conc %>% 
  group_by(element, srce_num) %>% 
  summarize(source_mean = mean(srce, na.rm = TRUE),
            MeanConc = mean(concentration, na.rm = TRUE)) %>% 
  ungroup() %>% 
  group_by(element) %>% 
  mutate(
    total = sum(source_mean),
    source_percent = round((source_mean/total)*100, digits = 2)) %>% 
  filter(element != "pm25")

# choose color palette
Color.list <- glasbey(26)

# make plot
conc_levels_bar <- 
  SA_proportion_full %>% 
  mutate(
    source_mean = ifelse(
      element %in% 
        c("OC", 
          "nitrate", 
          "ammonium", 
          "sulfur", 
          "EC"), 
      source_mean/10, 
      source_mean),
    group = ifelse(
      element %in% 
        c("OC", 
          "nitrate", 
          "ammonium", 
          "sulfur", 
          "EC"), 
      1, 0),
    element = as_factor(element),
    element = fct_relevel(element,
                          "ammonium",
                          "nitrate",
                          "EC",
                          "OC",
                          "sulfur",
                          "aluminum",
                          "arsenic",
                          "barium",
                          "bromine",
                          "cadmium",
                          "calcium",
                          "chlorine",
                          "chromium",
                          "copper",
                          "iron",
                          "lead",
                          "magnesium",
                          "manganese",
                          "nickel",
                          "potassium",
                          "selenium",
                          "silicon",
                          "sodium",
                          "titanium",
                          "vanadium",
                          "zinc"),
    element = fct_recode(element, 
                         "ammonium*" = "ammonium",
                         "nitrate*" = "nitrate",
                         "EC*" = "EC",
                         "OC*" = "OC",
                         "sulfur*" = "sulfur"),
    srce_num = as_factor(srce_num),
    srce_num = fct_recode(srce_num,
                          "Nitrate" = "1",
                          "Salt" = "2",
                          "Crustal Dust" = "3",
                          "Regional/Secondary" = "4",
                          "Traffic and Road Dust" = "5",
                          "Industrial" = "6")) %>%
  ggplot(aes(x = element, y = source_mean, group = group)) + 
  geom_col(aes(color = element, fill = element)) +
  geom_hline(yintercept = 0, size = 0.2) + 
  facet_wrap(. ~ srce_num) +
  theme_bw() +  
  theme(axis.text.x = element_blank(),
        axis.title = element_text(size = 16, face = "bold"),
        legend.text = element_text(size = 16, face = "bold"),
        legend.title = element_blank(),
        strip.text = element_text(face = "bold", size = 12)) +
  labs(x = expression("PM"[2.5]~"Constituents"),
       y = expression(paste("Predicted Average Concentration (", mu, "g/", m^3, ")*"))) +
  scale_color_manual(values = Color.list) +
  scale_fill_manual(values = Color.list)



# calculate proportion of each element coming from each source 
# (negative values from SA_Varimax converted to 0%)
SA_proportion_long <- source_conc %>% 
  group_by(element, srce_num) %>% 
  summarize(source_mean = mean(srce, na.rm = TRUE),
            MeanConc = mean(concentration, na.rm = TRUE)) %>% 
  mutate(
    source_mean = if_else(source_mean < 0, 0, source_mean)) %>% 
  ungroup() %>% 
  group_by(element) %>% 
  mutate(
    total = sum(source_mean),
    source_percent = round((source_mean/total)*100, digits = 2))

```

## Source proportions
```{r echo=FALSE}
# make the table nicer to look at

source_pie <- SA_proportion_long %>% 
  relocate(element, srce_num, source_percent) %>% 
  mutate(srce_num = plyr::mapvalues(srce_num,
           from = c("1", "2", "3", "4", "5", "6"), 
           to = c("Nitrate", "Salt", "Crustal Dust", "Secondary Sulfate", "Traffic + Road Dust", "Industrial"))) %>% 
  filter(element == "pm25") %>% 
  ggplot(aes(x = "", y = source_percent, fill = srce_num)) +
    geom_bar(stat = "identity", width = 1) +
    coord_polar("y", start = 0) +
    scale_fill_brewer(palette = "Spectral") +
    theme_void() +
    theme(legend.title = element_blank())

source_pie

vals <- c(48.9, 21.4, 15.9, 10.3, 1.96, 1.49)
val_names <- c("Secondary Sulfate (48.9%)", "Traffic and Road Dust (21.4%)", "Crustal Dust (15.9%)", "Nitrate (10.3%)", "Industrial (1.49%)", "Salt (1.96%)")
names(vals) <- val_names

waffle::waffle(vals, rows = 6, size = 0.6, colors = c("#73D055FF", "#FDE725FF", "#482677FF", "#287D8EFF","#404788FF", "#29AF7FFF")
  ) +
  theme(legend.text = element_text(face = "bold", size = 20))

SA_proportion <- SA_proportion_long %>% 
  relocate(element, srce_num, source_percent) %>% 
  filter(element != "pm25") %>% 
  pivot_wider(element:source_percent,
              names_from = srce_num,
              names_prefix = "source_",
              values_from = source_percent)

```

## Bar graph of the above proportions
```{r echo=FALSE}
# bar graph showing proportion of each element coming from each source
SA_bar <- SA_proportion_long %>% 
  ungroup() %>% 
  ggplot(aes(x = element, y = source_mean, fill = srce_num)) +
  geom_col(position = "fill") +
  theme(axis.text.x = element_text(angle = 90)) +
  ylab("") +
  xlab("") +
  scale_fill_brewer(palette = "Spectral")

SA_bar
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
# source contributions to total pm2.5
pm25 <- source_conc %>% 
  filter(element == 'pm25') %>% 
  select(-element) %>% 
  pivot_wider(names_from = srce_num,
              names_prefix = "source_",
              values_from = srce)


# write_csv(pm25, "./experimental_results/apca_scores.csv")

```
