---
title: "APCA"
author: "Rachel Tao"
date: "1/2/2021"
output: github_document
---

```{r setup, include=FALSE}
pacman::p_load(tidyverse, lubridate, ggmap, ggcorrplot, gplots, GGally, knitr, patchwork, psych, factoextra)

knitr::opts_chunk$set(
	fig.asp = 0.6,
	fig.width = 6,
	out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

```

Load in data

```{r echo=FALSE}
# dataset should include date, total pm2.5, and pm2.5 components

conc <- read_csv("./data/concentrations9910.csv") %>% 
  mutate(
    date = as.Date(date, "%m/%d/%Y"))

# multiply all concentrations by 1000 for units
i <- c(2:22)
conc[ , i] <- apply(conc[ , i], 2,
                    function(x)
                      x*1000)

```

Means, correlations and variances

```{r}
conc_corr <- 
  conc %>% 
  select(-date)

corr <- 
  summarize(conc_corr, corr = cor(conc_corr, use = "pairwise.complete.obs"))

conc_summarize <- 
  conc_corr %>% 
  pivot_longer(everything(),
               names_to = "element",
               values_to = "concentration") %>% 
  group_by(element)

means <- conc_summarize %>% 
  summarize(mean = mean(concentration, na.rm = TRUE)) %>% 
  arrange(element)

vars <- conc_summarize %>% 
  summarize(variance = var(concentration, na.rm = TRUE)) %>% 
  arrange(element)

```

Scaled factor analysis - must provide number of factors here

```{r}
# set mean = 0, SD = 1
# remove potassium so that 4th of July fireworks do not alter overall results
zspec_pm25 <- 
  conc %>% 
  select(-date, -pm25, -K) %>% 
  scale(center = TRUE, scale = TRUE)

# factor analysis - must provide number of factors!
factors <- 
  fa(zspec_pm25,
     nfactors = 6, max.iter = 500,
     rotate = "varimax",
     scores = "regression",
     SMC  = TRUE,
     fm = "pa")

# calculate factor scores, loadings, and weights (aka standardized scoring coefficients)
fact <- as_tibble(factors$scores)
loadings <- factors$loadings
factor_stat <- factors$weights %>% 
  as_tibble(rownames = NA) %>% 
  rownames_to_column("element") %>% 
  arrange(element)

```

calculate absolute factor scores for each day (Equations 5-7 in statistical methods of Kavouras 2001)

```{r}
# join standardized scoring coefficients with means and variances
coeff <- left_join(means, vars, by = "element") %>% 
  left_join(factor_stat, by = "element")

# calculate minimum values to find absolute factor scores (equations 5 and 6)
min <- coeff %>% 
  pivot_longer(contains("PA"),
               names_to = "factor_num",
               names_prefix = "PA",
               values_to = "factor_coeff") %>% 
  mutate(minimum = factor_coeff*(-mean/(variance^(1/2)))) %>% 
  select(element, factor_num, minimum) %>% 
  group_by(factor_num) %>% 
  summarize(sum = sum(minimum, na.rm = TRUE)) %>% 
  pivot_wider(factor_num:sum,
              names_from = factor_num,
              names_prefix = "minimum_",
              values_from = sum)

# reattach factor scores to dates and calculate absolute factor scores for each day
# (equation 7)
a <- conc %>% 
  select(date) %>% 
  cbind(fact) %>% 
  pivot_longer(contains("PA"),
               names_to = "factor_num",
               names_prefix = "PA",
               values_to = "factor_score") %>% 
  mutate(factor_num = as.numeric(factor_num)) %>% 
  cbind(min) %>% 
  pivot_longer(contains("minimum"),
               names_to = "min_num",
               names_prefix = "minimum_",
               values_to = "minimum") %>% 
  mutate(min_num = as.numeric(min_num)) %>% 
  filter(factor_num == min_num) %>% 
  mutate(factor_score = factor_score - minimum) %>% 
  select(-min_num, -minimum)

```

factor means

```{r}
# this is a check - not necessary for next steps
factor_means <- a %>% 
  select(factor_num, factor_score) %>% 
  group_by(factor_num) %>% 
  summarize(mean_score = mean(factor_score, na.rm = TRUE))

```

Regress each element on factors to see source contribution

```{r}
# define 'regress' as regression function with y as dependent variable and x as independent variable(s)
# return tidy output for beta values and r-squared value
regress = function(y) {
  
  z = lm(y ~ x)
  
  r.squared = broom::glance(z) %>% select(r.squared)

  betas = broom::tidy(z) %>% pivot_wider(term:estimate,
                                         names_from = "term",
                                         values_from = "estimate")
  
  return(cbind(r.squared, betas))
  
}

# independent variables for regression will be factors from factor analysis. define these as 'x'
x <- a %>% 
  pivot_wider(everything(),
              names_from = factor_num,
              names_prefix = "factor_",
              values_from = factor_score) %>% 
  select(contains("factor")) %>% as.matrix()

# dependent variable for regression will be actual element concentration. define this as y.
y <- conc %>% select(-date)

# repeat 'regress' for each element using 'x' as factors and 'y' as concentrations)
regress <- map(y, regress) %>% 
  enframe(name = "element") %>% 
  unnest(c(value)) %>% 
  janitor::clean_names() %>% 
  rename_at(vars(contains("factor")), funs(str_replace(., "xfactor", "beta")))

# dataset storing r-squred values
r <- regress %>% 
  select(element, r_squared)

# dataset storing beta values
betas <- regress %>% 
  select(element, starts_with("beta")) %>% 
  nest(betas = starts_with("beta"))

```

calculate source contributions

```{r}
# combine betas with factor scores in a list 
sources = lapply(betas$betas, cbind, a)

# in the above step, element names are automatically removed. rename list elements
names(sources) <- betas$element

# calculate source contributions by multiplying absolute factor scores by beta values
sources <- lapply(sources, as_tibble) %>% 
  bind_rows(.id = "element") %>%
  pivot_longer(starts_with("beta"),
               names_to = "beta_num",
               names_prefix = "beta_",
               values_to = "beta_value") %>% 
  mutate(
    beta_num = as.numeric(beta_num)
  ) %>% 
  filter(factor_num == beta_num) %>% 
  mutate(
    srce = factor_score*beta_value
  ) %>% 
  mutate(srce_num = as.character(beta_num)) %>% 
  select(element, date, srce_num, srce) 

conc_long <- conc %>% 
  pivot_longer(!date,
               names_to = "element",
               values_to = "concentration") %>% 
  arrange(element)

source_conc <- left_join(sources, conc_long, by = c('element', 'date'))

# calculate mean source contributions for each element (is this equivalent to laodings?)
mean_source <- source_conc %>% 
  group_by(element, srce_num) %>% 
  summarize(source_mean = mean(srce, na.rm = TRUE),
            MeanConc = mean(concentration, na.rm = TRUE)) %>% 
  pivot_wider(names_from = "srce_num",
              names_prefix = "source_",
              values_from = "source_mean")

```

### Final datasets:
 1) original loadings from factor analysis
 2) mean source contributions for each element, mean predicted concentration, percent error
 3) source contributions to total pm2.5

```{r}
# results of factor analysis, including loadings 
factors

# calculate mean predicted concentration and percent error
SA_Varimax <- mean_source %>% 
  ungroup(element) %>% 
  left_join(r, by = "element") %>% 
  mutate(
    PredConc = reduce(select(., contains("source")), `+`),
    Pct_error = (PredConc - MeanConc)/MeanConc
  )

# source contributions to total pm2.5
pm25 <- source_conc %>% 
  filter(element == 'pm25') %>% 
  select(-element) %>% 
  pivot_wider(names_from = srce_num,
              names_prefix = "source_",
              values_from = srce)

```
