---
title: "Explore APCA"
author: "Rachel Tao"
date: "3/14/2021"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

pacman::p_load(tidyverse, lubridate, ggmap, ggcorrplot, ggplot2, gplots, ggfortify, ggrepel, gridExtra, GPArotation, GGally, knitr, patchwork, psych, factoextra, RColorBrewer, kableExtra, pals, splines)

knitr::opts_chunk$set(
	echo = FALSE,
	fig.asp = 0.6,
	fig.width = 6,
	warning = FALSE,
	out.width = "90%"
)
theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis")

scale_colour_discrete = scale_colour_brewer(palette = "Spectral")
scale_fill_discrete = scale_fill_brewer(palette = "Spectral")

```

APCA cores were calcualted for years 2007-2015 using a 6-factor solution

```{r}
# Load in data

apca_scores_07 <- read_csv("./data/apca_scores_07.csv")

```

## general info

```{r}
summary(apca_scores_07, round(digits = 2))

summarize(apca_scores_07,
          mean = mean(concentration, na.rm = TRUE),
          sd = sd(concentration, na.rm = TRUE))
# what do I do about negative scores? keep them
# why are there 112 missings?
```

## time trends including 2007

```{r}
yr_num <- length(unique(year(apca_scores_07$date)))

ggplot(apca_scores_07, aes(x = date, y = concentration)) +
  geom_smooth(color = "red", method = lm, formula = y ~ ns(x, df = 1*yr_num)) +
  geom_smooth(color = "blue", method = lm, formula = y ~ ns(x, df = 5*yr_num))

# road dust?
ggplot(apca_scores_07, aes(x = date, y = source_1)) +
  geom_smooth(color = "red", method = lm, formula = y ~ ns(x, df = 1*yr_num)) +
  geom_smooth(color = "blue", method = lm, formula = y ~ ns(x, df = 5*yr_num))

# salt
ggplot(apca_scores_07, aes(x = date, y = source_2)) +
  geom_smooth(color = "red", method = lm, formula = y ~ ns(x, df = 1*yr_num)) +
  geom_smooth(color = "blue", method = lm, formula = y ~ ns(x, df = 5*yr_num))

# crustal dust
ggplot(apca_scores_07, aes(x = date, y = source_3)) +
  geom_smooth(color = "red", method = lm, formula = y ~ ns(x, df = 1*yr_num)) +
  geom_smooth(color = "blue", method = lm, formula = y ~ ns(x, df = 5*yr_num))

# regional/secondary
ggplot(apca_scores_07, aes(x = date, y = source_4)) +
  geom_smooth(color = "red", method = lm, formula = y ~ ns(x, df = 1*yr_num)) +
  geom_smooth(color = "blue", method = lm, formula = y ~ ns(x, df = 5*yr_num))

# traffic?
ggplot(apca_scores_07, aes(x = date, y = source_5)) +
  geom_smooth(color = "red", method = lm, formula = y ~ ns(x, df = 1*yr_num)) +
  geom_smooth(color = "blue", method = lm, formula = y ~ ns(x, df = 5*yr_num))

# industrial
ggplot(apca_scores_07, aes(x = date, y = source_6)) +
  geom_smooth(color = "red", method = lm, formula = y ~ ns(x, df = 1*yr_num)) +
  geom_smooth(color = "blue", method = lm, formula = y ~ ns(x, df = 5*yr_num))

```

# day of week patterns including 2007

```{r}

apca_scores_07 <- apca_scores_07 %>% 
  mutate(
    day_of_week = as.character(wday(date, label = TRUE)),
    day_of_week = factor(day_of_week,
                         levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat"))
    )

day_of_week_summary <- apca_scores_07 %>% 
  pivot_longer(source_1:source_6,
               names_to = "source",
               values_to = "score") %>% 
  group_by(day_of_week, source)

ggplot(day_of_week_summary, aes(day_of_week, score)) +  
  geom_bar(stat = "identity") + 
  facet_grid(source~., scales = "free")

```


Correlation Matrix

```{r}
apca_scores_07 %>% 
  select(-date, -day_of_week) %>% 
  rename(
    "Total PM" = concentration,
    "Nitrate" = source_1,
    "Salt" = source_2,
    "Crustal" = source_3,
    "Secondary" = source_4,
    "Traffic" = source_5,
    "Industrial" = source_6
    ) %>% 
  ggcorr(method = c("pairwise.complete.obs", "pearson"), limits = qFALSE,
       label = TRUE, label_size = 3, label_alpha = TRUE, label_round = 2,
       hjust = 1, size = 5, color = "grey50", layout.exp = 4)

```

Let's say 5 is spark-ignition traffic and 1 is diesel + road dust. Why would traffic be correlated with industrial? Could diesel + road dust also include construction, as the trucks used for construction may be diesel engines?